A New Score for Adaptive Tests\\in Bayesian and Credal Networks

### 2021-05-12 paper.tex
More stable version

\usepackage[]{algorithm2e}
\usepackage[]{algorithmic}
\usepackage{bm,hyperref,booktabs,tikz,pgf,pgfplots,amsmath}

@@

%\pgfplotsset{compat=1.13}
%\usepackage{pgfkeys}
%\usepackage{graphicx}
%xcolor} 
%amsmath} 
%amstext} 
%amssymb} 
%amsfonts} 
%\usepackage{mathptmx} 
%\usepackage[scaled=.90]{helvet} 
%\usepackage{courier} 
%\usepackage{textcomp} 
%\usepackage{array}
%\usepackage{varwidth}
%\usepackage[american]{babel}
%usepackage{mathtools} % fixes and extends amsmath
%mathtoolsset{mathic}
%\allowdisplaybreaks
%\usepackage{mathptmx}
%\usepackage{tabu}
%%\usepackage{breqn}
%\usepackage{algorithm,algorithmic,bm}
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%% Suggested packages, as needed:
%\usepackage{booktabs} % commands to create good-looking tables
%\usepackage{tikz} % nice language for creating drawings and diagrams - required for 
%drawing custom shapes
%\usetikzlibrary{%
%	arrows,%
%	calc,%
%	fit,%
%	patterns,%
%	plotmarks,%
%	shapes.geometric,%
%	shapes.arrows,%
%	shapes.callouts,%
	% shapes.multipart,%
	% shapes.gates.logic.US,%
	% shapes.gates.logic.IEC,%
	% er,%
	% automata,%
%	topaths,%
%	trees,%
%	matrix,%
%	% calendar,%
%	folding,%
%	fadings,%
%	positioning,%
%	scopes,%
%	decorations.fractals,%
%	decorations.shapes,%
%	decorations.text,%
%	decorations.pathmorphing,%
%	decorations.pathreplacing,%
%	decorations.footprints,%
%	decorations.markings,%
%	shadows}
%\usepackage{trace-pgfkeys}

@@

\title{Approximating Information-Theoretic Scores\\for Computer Adaptive Tests\\in Bayesian and Credal Networks}
%\titlerunning{Abbreviated paper title}

@@

%\authorrunning{F. Author et al.}

@@

A test is adaptive when the sequence and the number of questions is tuned on the basis of the estimated skills of the test taker. Graphical models such as Bayesian networks allow to model the uncertainty about the questions and the skills in an highly interpretable fashion, especially when coping with multiple skills. Yet, the quantification of the uncertainty in the question/skill relations can be better quantified by interval-valued probabilistic values in the model. This might make inference considerably harder in the underlying model, this being especially the case for the information theoretic quantities generally used as a score to drive the adaptation process.
Here we present an alternative score based on the mode of the posterior probabilities of the model. We show how this makes considerably simpler the evaluation of the score in the credal case. Numerical tests are promising, we show how the performance are similar in the Bayesian network case, while xxx.
%considered. 
%We propose an alternative 
%We present.
%with xxx. Interval valued quantification
%We suggest the use of credal networks, a generalization of Bayesian networks based on sets of probability mass functions, to implement adaptive tests exploiting the knowledge of the test developer instead of training on databases of answers. Compared to Bayesian networks, these models might offer higher expressiveness and hence a more reliable modeling of the qualitative expert knowledge. The counterpart is a less straightforward identification of the 
%information-theoretic measure controlling the question-selection and the test-stopping criteria. We elaborate on these issues and propose a sound and computationally feasible procedure. Validation against a Bayesian-network approach on a benchmark about German language proficiency assessments suggests that credal networks can be reliable in assessing the student level and  effective in reducing the number of questions required to do it.
%150-200 words - 12 pages
\keywords{First keyword  \and Second keyword \and Another keyword.}

@@

\section{A Toy Example}
As a first, demonstrative, example of our ideas consider a model characterized by a single (Boolean) Skill $S$ and a set of questions $\bm{Q}$ picked from a set of templates. Let us first focus on a Bayesian, i.e., precise, case. The prior specification $P(S)$ is achieve by a single parameter $p \in [0,1]$ such that $P(S=1)=p$. For a generic question $Q$, which is also assumed to be Boolean, a complete specification of the CPT $P(Q|S)$ is obtained by the two parameters $p_0,p_1\in [0,1]$ where $P(Q=1|S=0)=p_0$ and $P(Q=1|S=1)=p_1$. In other wrods $p_0$ denote the probability of providing a right answer to the question for a student who does not have the considered skill and $p_1$ for a student who have that skill. An obvious constraint is $p_1>p_0$, i.e., having the skill makes your probability of providing a right answer higher. The inequality should be strict, otherwise the two probabilities are equal, and this means that the answer is irrelevant for the skill. Although parameters $p_1$ and $p_0$ already give a complete parametrization, we perform the linear transformation
$\lambda:=1-\frac{1}{2}(p_0+p_1)$ and $\delta=(p_1-p_0)$. The first quantity is the arithmetic average of giving a wrong answer and we might indent it as the \emph{level} of the question, this mining that the higher is $\lambda$ the more unlikely is to receive a correct answer is the students have the same probability of having and not having the skill. Regarding $\delta$ we might intend it as the (average) discriminative power of a question.
Note that the inverse relations give $p_1=1-\lambda+\frac{\delta}{2}$
and $p_0=1-\lambda-\frac{\delta}{2}$. Where constraint $p_1>p_0$ is just $\delta>0$ and any other value in $[0,1]$ for both parameters is valid.

In our toy example, we consider nine possible templates, made of triads of templates with the same $\delta$ and same $\lambda$. The reference values $[0.4,0.5,0.6]$ are used.

\section{Testing Algorithms}
The typical goal of a test or exam is to evaluate the knowledge level of a student or examinee $\sigma$ on the basis of her answers to a number of questions or exercises. Let $\bm{Q}$ denote a repository of questions available to the instructor. The order and the number of questions from $\bm{Q}$ asked to $\sigma$ might not be fixed. We call \emph{testing algorithm} (TA) a procedure taking care of the selection of the sequence of questions asked to the student and able to decide when the test should stop. A standard TA pseudocode is in Algorithm \ref{alg:ta}, where $\bm{e}$ denotes the array of the answers collected from the student $\sigma$.

\begin{algorithm}[H]
    \SetAlgoLined
    \KwResult{Write here the result }
    initialization\;
    \While{While condition}{
        instructions\;
        \eIf{condition}{
            instructions1\;
            instructions2\;
        }{
            instructions3\;
        }
    }
    \caption{How to write algorithms}
\end{algorithm}

\begin{algorithm}[htp]
    \begin{algorithmic}[1]
    \STATE $\bm{e}=\emptyset$
        \WHILE{ {\bf not} $\tt{Stopping}(\bm{e})$}
        \STATE $Q^* \gets \tt{PickQuestion}(\bm{Q},\bm{e})$
        \STATE $q^* \gets {\tt{Answer}}(Q^*,\sigma)$
        \STATE $\bm{e} \gets \bm{e} \cup \{ Q^*=q^* \}$
        \STATE $\bm{Q} \gets \bm{Q} \setminus \{ Q^*\}$
        \ENDWHILE
        \STATE {\bf return} $\tt{Evaluate}(\bm{e})$
    \end{algorithmic}\label{alg:ta}
\caption{Testing algorithm. Student profile $\sigma$ and question repository $\bm{Q}$ are inputs.}
\end{algorithm}

The Boolean function ${\tt Stopping}$ decides whether or not the test should end, this choice being possibly based on the previous answers as collected in $\bm{e}$. Trivial examples of such stopping rules might be based on the number of questions asked to the student (${\tt Stopping}(\bm{e})=1$ if and only if $|\bm{e}|>n$) or on the number of correct answers provided that a maximum number of questions is not exceeded. Function ${\tt PickQuestion}$ is instead in charge of deciding which question should be asked to the student from the repository $\bm{Q}$. A TA is called \emph{adaptive} when this function is based on the previous answers $\bm{e}$. Trivial non-adaptive strategies might consist in randomly picking an element of $\bm{Q}$ or following a fixed order. Function ${\tt Answer}$ is simply collecting (or simulating) the answer of the student $\sigma$ to a particular question $Q$. In our assumptions, this answer is not affected by the previous answers to other questions.\footnote{Generalized setups where the quality of the student answer is affected by the previous answers will be discussed at the end of the paper. This might include a \emph{fatigue} model negatively affecting the quality of the answers when many questions have been already answered as well as the presence of \emph{revealing} questions that might improve the quality of other answers.}

Finally, the ${\tt Evaluate}$ function decides the overall judgement for the test (e.g., a numerical grade or a pass/fail Boolean) on the basis of all the answers collected after the test termination. Trivial examples of such function are the percentage of correct answers or a Boolean that is true when a sufficient percentage of correct answers has been provided. Note also that in our assumptions the TA is \emph{exchangeable}, i.e., the stopping rule, the question finder and the evaluation function are invariant with respect to permutations of $\bm{e}$. In other words, the would pick the same next question and the same evaluation for two students who provided the same list of answers in two different orders and similarly would decide to stop or not to stop in the same way.

The typical goal of a TA is to achieve a reliable evaluation of the student $\sigma$. As each answer is assumed to improve such quality, asking all the questions, no matter their order because of the exchangeability assumption, would be an obvious choice. Yet, asking all the questions in the repository might be impractical (e.g., because of time limitations) in some cases or just provide an unnecessary burden to the students. The goal of a good TA is therefore to trade off the accuracy of the evaluation and the number of questions.\footnote{In some generalized setups, other elements such as the \emph{serendipity} to avoid boring sequences of questions might be also considered.}

Item response theory is a popular *** Something about IRT here ***. Most of these approaches are based an a unidimensional representation of the student knowledge level. Proposals to cope with multidimensional IRTs have been considered, but this results critical when coping with dependencies between them. For this reason probabilistic modelling can be considered. This is the topic of the next section.

\section{Bayesian Adaptive Testing Algorithms}
In the setup of algorithm.

Let us introduce the discussion by means of a simple example modelling the knowledge level of a student in a particular domain. In order to evaluate such level a number of questions can be asked to the student. We describe the knowledge level (or skill) of the student by means of a single variable $S$. We assume $S$ to be a discrete, ordinal, variable, whose $m$ states $\mathcal{S}:=(s^{(0)},\ldots,s^{(m-1)})$ can be intended as increasing knowledge levels in that domain. Regarding the questions, we initially describe them as Boolean variables, whose true state correspond to a correct answer to the question, while no distinctions between the possible wrong answers are considered.\footnote{Dedicated discussion about abstention.} Let $Q$ with $\mathcal{Q}:=(q,\overline{q})$ denote a generic question we ask to the student. A repository of $n$ questions $\bm{Q}$ is assumed to be available. A \emph{testing} algorithm is a procedure deciding which question and how many questions from $\bm{Q}$ should be asked to the student, on the basis of her previous answers. Algorithm X provides an abstract description of xxx.

	%
	%
	\section{Brainstorming}

	\subsection{Linear Programs for Credal Mode}

	Given the Boolean question $Q$ and the Boolean skill $S$, we use the score:

	\begin{equation}
		\sigma_Q := \max_{P\in K} \bigl[ 
		\max \{ P(S=0,Q=0) , P(S=1,Q=0) \}  + \max \{ P(S=0,Q=1),P(S=1,Q=1)\}
		\bigr]
	\end{equation}

	\subsection*{Pseudocode}
	Let us first specify the inputs:
	\begin{itemize}
		\item Let $\bm{S}:=(S_1,\ldots,S_N)$ denote the set of skills in the model. We 
		describe a \emph{student} as a joint observation $\bm{s}$ of the skills.
		\item Let $M$ denote the network \emph{skeleton}, i.e., the model over the skills only
		\item A question $Q$ is a binary variables with states right and wrong
		\item Each question depends on a single skill only, let $\mathrm{Pa}(Q)$ denote this 
		skill
		\item Let $\tau(S)$ denote the number of \emph{templates} for the CPT of a question 
		associated with $S$, $\{P(Q|S)\}_{t=1}^{\tau(S)}$ 
		\item Let $n_q$ be the maximum number of questions we might ask for each skill
	\end{itemize}
	The algorithm works as follows.
	\begin{enumerate}
		\item For eachLet $\bm{S}:=(S_1,\ldots,S_N)$ denote the set of skills in the model
		\item Let $M$ denote the network \emph{skeleton}, i.e., the model over the skills only
		\item A question $Q$ is a binary variables with states right and wrong
		\item Each question depends on a single skill only, let $\mathrm{Pa}(Q)$ denote this 
		skill
		\item Let $\tau(S)$ denote the number of \emph{templates} for the CPT of a question 
		associated with $S$, $\{P(Q|S)\}_{t=1}^{\tau(S)}$ 
		\item Let $n_q$ be the maximum number of questions we might ask for each skill
	\end{enumerate}

%	\begin{algorithm}[htp!]
%		\caption{Single skill, precise model\\
%			a model (BN, skeleton) $M$ over the skills only
%			\\a dataset of $n$ students $\{s^i\}_{i=1}^n$,\\a dataset of $m$ questions 
%			$\mathcal{Q}:=\{Q^j\}_{j=1}^m$,\\
%			the CPTs of the questions in $\mathcal{Q}$, i.e., $\{P(Q^j|S)\}_{j=1}^m$\\
%			the student answer models $\{\tilde{P}(Q^j|S)\}$\\
%			an entropy threshold $h^*$,\\
%			a score function for questions $\gamma$
%			\label{alg:simple}}
%		\begin{algorithmic}[1]
%			\FOR {$i\gets\{1,\ldots,n\}$}
%			\STATE $(M',\mathcal{Q}',\mathcal{Q}'',\bm{e}) \gets 
%			(M,\mathcal{Q},\emptyset,\emptyset)$
%			\WHILE{ $h > h^* \mathrm{\bf and}\, \mathcal{Q}' \neq \mathcal{Q}$}
%			\STATE $Q^* \gets \arg\max_{Q \in \mathcal{Q}'}\gamma(M',Q,\bm{e})$
%			\STATE{$\mathcal{Q}'\gets \mathcal{Q}' \setminus \{ Q^*\}$}
%			\STATE{$\mathcal{Q}''\gets \mathcal{Q}'' \cup \{ Q^*\}$}
%			\STATE{$q^* \gets \mathrm{sampling} \sim \tilde{P}(Q^*|s^i)$}
%			\STATE $M' \gets \mathrm{addChild}(M',Q,S,P(Q|S))$
%			\STATE $\bm{e} \gets \bm{e} \cup \{ Q^*=q^* \}$
%			\STATE $h \gets H[P_{M'}(S|\bm{e})]$
%			\ENDWHILE
%			\STATE $\sigma_i \gets \arg\max_{s\in\mathcal{S}} P_{M'}(s|\bm{e})$
%			\ENDFOR
%			\STATE {\bf return} $(\sigma_1,\ldots,\sigma_n)$
%		\end{algorithmic}
%	\end{algorithm}
	Three possible scores:
	\begin{eqnarray}
		\gamma(Q,\bm{e}) := -H[S|Q,\bm{e}] = -\sum_q P(q) \cdot H[S|\bm{e},q]&&\\
		\gamma'(Q,\bm{e}) :=  -\sum_q P(q) \cdot \max_{s\in\mathcal{S}}  P[s|\bm{e},q]&&\\
		\gamma''(Q,\bm{e}) :=  |P(Q=1|\bm{e})-0.5|&&
	\end{eqnarray}

	\textcolor{red}{Maybe, instead of (3) I would use a known measure of qualitative 
	variation (\url{https://en.wikipedia.org/wiki/Qualitative_variation}) which extend easily to 
	non binary answers, e.g., the variation ratio ($1-\max_q(P(q))$) or the M1 Gibb's index 
	also known as coefficient of unalikeability and somehow related to the variance (it is 
	the variance in the binary case): $M1 = 1-\sum_q P(q)^2$, or even the entropy, which is 
	not so difficult to compute for $P(Q)$. They are all equivalent in the binary case, as 
	they are all monotonic function of $\max_q(P(q))$ only. }

%	\begin{algorithm}[htp!]
%		\caption{Single skill, credal model\\a dataset of $n$ students $\{s^i\}_{i=1}^n$,\\a 
%		dataset of $m$ questions $\mathcal{Q}:=\{Q^j\}_{j=1}^m$,\\
%			the credal CPTs of the questions in $\mathcal{Q}$, i.e., 
%			$\{\overline{P}(Q^j|S)\}_{j=1}^m$\\
%			an upper entropy threshold $\overline{h}^*$,\\
%			a score function for questions $\sigma$
%			\label{alg:simple2}}
%		\begin{algorithmic}[1]
%			\FOR {$i\gets\{1,\ldots,n\}$}
%			\STATE $(\mathcal{Q}',\mathcal{Q}'',\bm{e}) \gets 
%			(\mathcal{Q},\emptyset,\emptyset)$
%			\WHILE{ $h > h^* \mathrm{\bf and}\, \mathcal{Q}' \neq \mathcal{Q}$}
%			\STATE $Q^* \gets \arg\max_{Q \in \mathcal{Q}'}\gamma(Q,\bm{e})$
%			\STATE{$\mathcal{Q}'\gets \mathcal{Q}' \setminus \{ Q^*\}$}
%			\STATE{$\mathcal{Q}''\gets \mathcal{Q}'' \cup \{ Q^*\}$}
%			\STATE{$q^* \gets \mathrm{sampling} \sim P(Q^*|s^i)$}
%			\STATE $\mathrm{addChild}(Q,S,P(Q|S))$
%			\STATE $\bm{e} \gets \bm{e} \cup \{ Q^*=q^* \}$
%			\STATE $h \gets H[P(S|\bm{e})]$
%			\ENDWHILE
%			\STATE $\sigma_i \gets \arg\max_{s\in\mathcal{S}} P(s|\bm{e})$
%			\ENDFOR
%			\STATE {\bf return} $(\sigma_1,\ldots,\sigma_n)$
%		\end{algorithmic}
%	\end{algorithm}
	Three possible scores:
	\begin{eqnarray}
		\gamma(Q,\bm{e}) := -H[S|Q,\bm{e}] = -\sum_q P(q) \cdot H[S|\bm{e},q]&&\\
		\gamma'(Q,\bm{e}) :=  -\sum_q P(q) \cdot \max_{s\in\mathcal{S}}  P[s|\bm{e},q]&&\\
		\gamma''(Q,\bm{e}) :=  |P(Q=1|\bm{e})-0.5|&&
	\end{eqnarray}

>>>>

\usepackage{algorithm,algorithmic,amsfonts,bm,hyperref,booktabs,tikz,pgf,pgfplots,amsmath,soul,subcaption,url,graphicx,amsmath,booktabs,booktabs,nicefrac,bm,colortbl,tikz,pgfplots,enumerate,xargs}

@@

\captionsetup{compatibility=false}
\usetikzlibrary{arrows,plotmarks,decorations.markings,trees,shapes}
\tikzstyle{nodo}=[ellipse,draw=black!100,fill=black!30,line width=.7pt,minimum width=1.2cm,minimum height=.7cm]
\tikzstyle{Qnodo}=[ellipse,draw=black!100,fill=black!10,line width=.7pt,minimum width=1.2cm,minimum height=.7cm]
\tikzstyle{arco}=[draw=black!80,line width=.7pt, postaction={decorate}, decoration={markings,mark=at position 1.0 with {\arrow[ draw=black!80,line width=.7pt]{>}}}]
\tikzstyle{decision} = [rectangle, draw, fill=black!100,text=white, text width=4.5em, text badly centered, node distance=3cm, minimum height=3em]
\tikzstyle{block} = [rectangle, draw, fill=blue!20, text width=5em, text centered, rounded corners, minimum height=3em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm, minimum height=2em]
\def\mystrut{\vphantom{hg}}
\pgfplotsset{legend image with text/.style={
legend image code/.code={%
\node[anchor=center] at (0.3cm,0cm) {#1};}},}
\pgfplotsset{compat=1.13}

@@

\title{A New Score for Adaptive Tests\\in Bayesian and Credal Networks}

@@

%\titlerunning{A New ScoreAbbreviated paper title}
\authorrunning{Antonucci et al.}

@@

A test is \emph{adaptive} when its sequence and number of questions is dynamically tuned on the basis of the estimated skills of the taker. Graphical models, such as Bayesian networks, are used for adaptive tests as they allow to model the uncertainty about the questions and the skills in an explainable fashion, especially when coping with multiple skills. A better elicitation of the uncertainty in the question/skills relations can be achieved by interval probabilities. This turns the model into a \emph{credal} network, thus making the inferential complexity of the queries required to select questions more demanding. This is especially the case for the information theoretic quantities used as \emph{scores} to drive the adaptive mechanism. We present an alternative family of scores, based on the mode of the posterior probabilities and hence easier to explain. This makes considerably simpler the evaluation in the credal case, without significantly affecting the quality of the adaptive process. Numerical tests on synthetic and real-world data are used to support this claim.
%12 pages
\keywords{Computer adaptive tests \and Information theory \and Credal networks \and Bayesian networks}

@@

\section{Introduction}\label{sec:intro}
A test or an exam can be naturally intended as a measurement process, with the questions acting as sensors measuring the skills of the test taker in a particular discipline. Such measurement is typically imperfect with the skills acting as latent variables whose actual values cannot be revealed in a perfectly reliable way. The role of the questions, whose answers are regarded instead as manifest variables, is to reduce the uncertainty about the latent skills. Following this perspective, probabilistic modelling is an obvious framework to describe tests. Consider for instance the example in Figure \ref{fig:minicat}, where a small Bayesian network is used to describe the probability of the test taker having understood how to multiply numbers. In such a framework making the test \emph{adaptive}, i.e., picking a next question on the basis of the current knowledge level of the test taker is also very natural. The expected information gain for the available questions might be naturally used to select the question leading to the more informative results about the skill level of the test taker. This might take place before the answer to the question on the basis of the expectations about the possible answers.

A critical point when coping with such approaches is to provide a realistic assessment for the probabilistic parameters associated with the modelling of the relations between the questions and the skills. Having to provide sharp numerical values for these probabilities might be difficult. As the skill is typically a latent quantity, complete data are not available for a statistical learning of the parameters and the elicitation should therefore demanded to experts (e.g., a teacher). Yet, it might be not obvious to express such a domain knowledge by single numbers and a more robust elicitation, such as a probability interval, might add realism and robustness to the modelling process \cite{hajek2012rationality}. With such generalized assessments of the parameters a Bayesian network simply becomes a \emph{credal} network \cite{antonucci2010d}. Yet, the counterpart of such increased realism is the higher computational complexity characterizing the inference in credal networks \cite{maua14jair}. This is especially the case when coping with information theoretic measures such a information gain, whose computation might lead to complex non-linear optimization tasks \cite{mangili2017b}. The goal of this paper is to investigate the potential of alternatives to the information-theoretic scores driving the question selection in adaptive tests based on directed graphical models, no matter whether these are Bayesian or credal networks. In particular, we consider a family of scores based on the (expected) mode of the posterior distributions over the skills. We show that, when coping with credal networks, the computation of these scores can be reduced to the simpler optimization tasks involving multi-linear optimizations for which powerful approximation schemes are already available \cite{antonucci2013a}. Moreover, we show that these scores benefit of better explainability properties, thus allowing for a more transparent process in the question selection.

\begin{figure}[htp!]
\centering
\begin{tikzpicture}[scale=1]
\node[] ()  at (-5,0.5) {\tiny $P(S=1|Q=1)=0.5$};
\node[] ()  at (-5,-2) {\tiny $P(S=1|Q=1)=0.5$};
\node[] ()  at (-5,-3) {\tiny $P(S=1|Q=1)=0.5$};
\node[] ()  at (3,0.5) {\tiny $P(S=1)=0.5$};
\node[] ()  at (0,-3) {\tiny $2x3=?$ \quad $2x3=?$\quad  $2x3=?$};
\node[] ()  at (0,-2) {\tiny $P(Q=1|S=1)=0.3$ \quad $0.4$\quad  $0.5$};
\node[] ()  at (0,-2.5) {\tiny $P(Q=1|S=0)=0.3$ \quad  $0.4$ \quad  $0.5$};
\node[nodo] (s)  at (0,0.5) {$S$};
\node[Qnodo] (q1)  at (-2.,-1) {$Q_1$};
\node[Qnodo] (q2)  at (0,-1) {$Q_2$};
\node[Qnodo] (q3)  at (2,-1) {$Q_3$};
\draw[arco] (s) -- (q1);
\draw[arco] (s) -- (q2);
\draw[arco] (s) -- (q3);
\end{tikzpicture}
\caption{Ad adaptive testing scheme to evaluate the multiplication (Boolean) skill $S$ by three (Boolean) questions. Probabilities are assessed by the Bayesian network on the right}
    \label{fig:minicat}
\end{figure}
*** espandere figura ***
\newpage 
The paper is organized as follows. A critical discussion about the existing work in this area is in Section \ref{sec:work}. The necessary background material is reviewed in Section \ref{sec:background}. The adaptive testing concepts are introduced in Section \ref{sec:cat} and specialized to graphical models in \ref{sec:bncat}. The technical part of the paper is in Section \ref{sec:mode}, where the new scores are discussed and specialized to the credal case, while the experiments are in Section \ref{sec:exp}. Conclusions and outlooks are in Section \ref{sec:conc}.

\section{Related Work}\label{sec:work}
Modelling a test as a process involving latent variables is a classical approach since the proposal of the classical \emph{item response theory}, that has been widely used even to implement adaptive sequences \cite{xxx}. Despite its success related to the ease of implementation and inference, IRT might be inadequate when coping with multiple latent skill, especially when these are dependent. For such cases, interpretable graphical models such as Bayesian networks have been resulted more effective [x]. The development of (computer) adaptive tests is a classical topic for AI \cite{xx}. Apart from graphical models, adaptive algorithms have been proposed within a genetic optimization framework. The idea of using Bayesian network in grading environment is not new. xxx. As already noticed in the previous section, the latent nature of the student skills might hinder the direct statistical learning of the parameters of such models. This naturally led some authors to consider missing data modelling. In particular the work of xxxx Vomlel related to EM algorithm in Bayesian networks is a notable example of this kind of approaches. Some other authors considered even more extreme directions involving missingness even about the answers such as \cite{bachrach2012grade}. This is also related to the notion of monotonicity \cite{plajner2020monotonicity}. The potential of these models in terms of explainability of the results has been emphasized in the work Work of Darwiche about same-probability problem. *** completare *** Bayesian networks (BNs) \cite{koller2009} have been used to model the knowledge driving such intelligent systems \cite{almond2015bayesian}. However, collecting large sets of reliable data in educational domains may be difficult and time consuming (e.g., a course with few students, or taught for the first time), and the quantification should be based on expert knowledge only. To elicit a Bayesian network, an expert might face questions like: ``\emph{which is the probability of a student with a particular knowledge level giving the right answer to a question?}''. Giving sharp probabilities for questions of this kind can be problematic for an expert, whose knowledge is mostly qualitative (e.g., ``\emph{a right answer is very unlikely}''). Fuzzy linguistic approaches represent a viable, non-numerical, way to address these issues \cite{badaracco2013fuzzy}. To stick within the probabilistic framework, verbal-numerical probability scales associated with sharp values \cite{renooij1999talking} or intervals \cite{walley1991statistical} have been also proposed. a generalization of BNs based on the imprecise probability theory \cite{walley1991statistical}, where local parameters are defined by set-valued probabilities. This simplifies the elicitation process and offers a more reliable handling of the related uncertainty. Moving from BNs to CNs implies two main issues: (i) numerical inferences will be interval-valued too, thus making debatable both the decision criterion \cite{troffaes} an the information measures \cite{klir1999uncertainty} to adopt; and (ii) inference tasks in CNs typically belongs to higher complexity classes than their Bayesian counterparts \cite{maua14jair}. Both these issues are addressed by defining a computationally feasible procedure based on CNs to be used for practical implementation of intelligent systems solely specified by expert knowledge. To the best of our knowledge this is the first attempt to perform e-testing with models of this kind.

We focus on the application of CNs to \emph{computer adaptive testing} (CAT), i.e., an approach to e-testing that adjusts the sequence and the number of questions to the ability level of the test taker. CATs have the potential to make the test an individualised experience that challenges and does not discourage the test takers, as most of the questions are near their ability levels. Building upon \emph{item response theory} \cite{hambleton1985item}, the common background underpinning CATs, graphical modeling (such as BNs and CNs) offers a powerful language for describing complex multivariate dependencies between skills and rich tasks. Several researchers have exploited the potential of BNs both in adaptive and non-adaptive educational assessment \cite{vomlel2004building,plajner2015}. These authors focus on applications for which data are available to learn the model parameters. We regard this point as a serious limitation, possibly hindering CATs adoption by many teachers and instructors. The results are promising: CAT based on CNs is effective in reducing the number of questions while maintaining a high accuracy in the evaluation and the approximations introduced do not compromise the procedure's effectiveness.

\section{Background on Bayesian and Credal Networks}\label{sec:background}
We denote variables by Latin uppercase letters, while using lowercase for their generic values, and calligraphic for the set of their possible values. Thus, $x \in \mathcal{X}$ is a possible value of $X$. Here only variables with a finite number of possible values are considered.\footnote{Adaptive approaches such as IRT \cite{xx} adopt continuous latent variables.}

\subsection{Bayesian Networks}
A probability mass function over $X$ is denoted as $P(X)$, while $P(x)$ is the probability assigned to state $x$. Given a function $f$ of $X$, its expectation with respect to $P(X)$ is $\mathbb{E}_P(f):=\sum_{x\in\mathcal{X}} P(x) \cdot f(x)$. The expectation of $-\log_k[P(X)]$ is called \emph{entropy} and denoted also as $H_P(X)$.\footnote{To cope with zero probabilities, we set $0 \cdot \log_k 0 = 0$.} We assume $k:=|\mathcal{X}|$ to have the maximum of the entropy, achieved for uniform PMFs, equal to one.

Given a joint PMF $P(X,Y)$, the marginal PMF $P(Y)$ is obtained by summing out the other variable, i.e., $P(y)=\sum_{x\in\mathcal{X}} P(x,y)$. Conditional PMFs such as $P(X|y)$ are similarly obtained by Bayes's rule, i.e., $P(x|y)=P(x,y)/P(y)$ provided that $P(y)>0$. Notation $P(X|Y):=\{P(X|y)\}_{y\in\mathcal{Y}}$ is used for such conditional probability table (CPT). The entropy of a conditional PMF is defined as in the unconditional case and denoted as $H_P(X|y)$. The expected entropy is a weighted average of conditional entropies, i.e., $H_P(X|Y):=\sum_{y\in\mathcal{Y}} H_P(X|y) P(y)$. If $P(x,y)=P(x) P(y)$ for each $x\in\mathcal{X}$ and $y\in\mathcal{Y}$, variables $X$ and $Y$ are stochastically independent. A conditional formulation can be also considered.

We assume the set of variables $\bm{X}:=(X_1,\ldots,X_n)$ to be in one-to-one correspondence with a directed acyclic graph $\mathcal{G}$. For each $X\in\bm{X}$, the parents of $X$ are called predecessor of $X$ according to $\mathcal{G}$ and denoted as $\mathrm{Pa}_X$. The graph $\mathcal{G}$ together with the collection of CPTs $\{P(X|\mathrm{Pa}_X)\}_{X\in\bm{X}}$ provides a Bayesian network (BN) specification \cite{koller2009}. Under the Markov condition, that states that every variable is conditionally independent of its non-descendants non-parents given its parents, a BN compactly defines a joint PMF $P(\bm{X})$ that factorizes as $P(\bm{x})=\prod_{X\in\bm{X}} P(x|\mathrm{pa}_X)$.

\subsection{Credal Sets and Credal Networks}
A set of PMFs over $X$ is denoted as $K(X)$ and called \emph{credal set} (CS). Expectations based on CSs are the bounds of the PMF expectations with respect to the CS. Thus $\underline{E}_K[f]:=\inf_{P(X)\in K(X)} \mathbb{E}_P[f]$ and similarly for the supremum $\mathbb{E}_K$. Expectations of events are in particular caller lower and upper probabilities and denotes as $\underline{P}$ and $\overline{P}$. Notation $K(X|y)$ is used for a set of conditional CSs, while $K(X|Y):=\{K(X|y)\}_{y\in\mathcal{Y}}$ is a credal CPT (CCPT).

Analogously to a BN, a credal network (CN) \cite{cozman2000}  is specified by graph $\mathcal{G}$ together with a family of CCPTs $\{K(X|\mathrm{Pa}_X)\}_{X\in\bm{X}}$. A CN defines a joint CS $K(\bm{X})$ obtained by the set of xxx.

\section{Testing Algorithms}\label{sec:cat}
A typical test aims at evaluating the knowledge level of a test taker $\sigma$ on the basis of her answers to a number of questions. Let $\bm{Q}$ denote a repository of questions available to the instructor. The order and the number of questions from $\bm{Q}$ asked to $\sigma$ might not be defined in advance. We call \emph{testing algorithm} (TA) a procedure taking care of the selection of the sequence of questions asked to the test taker, and to decide when the test stops. Algorithm \ref{alg:ta} depicts a general TA scheme, with $\bm{e}$ denoting the array of the answers collected from the test taker $\sigma$.

\begin{algorithm}[htp!]
\begin{algorithmic}[1]
\STATE $\bm{e}\gets\emptyset$
\WHILE{ {\bf not} $\tt{Stopping}(\bm{e})$}
\STATE $Q^* \gets \tt{Pick}(\bm{Q},\bm{e})$
\STATE $q^* \gets {\tt{Answer}}(Q^*,\sigma)$
\STATE $\bm{e} \gets \bm{e} \cup \{ Q^*=q^* \}$
\STATE $\bm{Q} \gets \bm{Q} \setminus \{ Q^*\}$
\ENDWHILE
\STATE {\bf return} $\tt{Evaluate}(\bm{e})$
\end{algorithmic}
\caption{General TA: given the profile $\sigma$ and repository $\bm{Q}$, an evaluation based on answers $\bm{e}$ is returned.\label{alg:ta}}
\end{algorithm}

Boolean function ${\tt Stopping}$ decides whether the test should end, this choice being possibly based on the previous answers in $\bm{e}$. Trivial stopping rules might be based on the number of questions asked to the test takes (${\tt Stopping}(\bm{e})=1$ if and only if $|\bm{e}|>n$) or on the number of correct answers provided that a maximum number of questions is not exceeded. Function ${\tt Pick}$ selects instead the question to be asked to the student from the repository $\bm{Q}$. A TA is called \emph{adaptive} when this function takes into account the previous answers $\bm{e}$. Trivial non-adaptive strategies might consist in randomly picking an element of $\bm{Q}$ or following a fixed order. Function ${\tt Answer}$ is simply collecting (or simulating) the answer of test taker $\sigma$ to a particular question $Q$. In our assumptions, this answer is not affected by the previous answers to other questions.\footnote{Generalized setups where the quality of the student answer is affected by the previous answers will be discussed at the end of the paper. This might include a \emph{fatigue} model negatively affecting the quality of the answers when many questions have been already answered as well as the presence of \emph{revealing} questions that might improve the quality of other answers.}

Finally, ${\tt Evaluate}$ is a function returning the overall judgement of the test (e.g., a numerical grade or a pass/fail Boolean) on the basis of all the answers collected after the test termination. Trivial examples of such functions are the percentage of correct answers or a Boolean that is true when a sufficient number of correct answers has been provided. Note also that in our assumptions the TA is \emph{exchangeable}, i.e., the stopping rule, the question finder and the evaluation function are invariant with respect to permutations in $\bm{e}$. In other words, the same next question, the same evaluation and the same stopping decision is produced for any two students, who provided the same list of answers in two different orders.

A typical TA is required to achieve a reliable evaluation of the student $\sigma$ after the answers $\bm{e}$. As each answer is individually assumed to improve such quality, asking all the questions, no matter in which  order because of the exchangeability assumption, is an obvious choice. Yet, this might be impractical (e.g., because of time limitations) or just provide an unnecessary burden to the test taker. The goal of a good TA is therefore to trade off the accuracy of the evaluation and the number of questions.\footnote{In some generalized setups, other elements such as a \emph{serendipity} in choice in order to avoid tedious sequences of questions might be also considered \cite{xx}.}
